---
title: "Problem Set: Vectors, Strings, Factors, and Dates"
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(nycflights13)
library(lubridate)
```

------------------------------------------------------------------------

## Question 1 — Exploring the Star Wars Universe

The `dplyr::starwars` dataset (available as soon as you load `tidyverse`) contains physical and biographical information on characters from the Star Wars films. Start by running `View(starwars)` and ?starwars to get oriented.

The code below stores the characters with blue eyes in the "blue_eyes" dataframe, and the characters with red eyes in the "red_eyes" dataframe. There is one character with both blue and red eyes. Use a logical vector approach again with str_detect to find out which one. Hint: one way to do it is with the logical & operator, which returns all rows for which multiple criteria are met

```{r}
# Characters with "blue" in eye_color
blue_eyes <- starwars |>
  filter(str_detect(eye_color, "blue"))
blue_eyes

# Characters with "brown" in eye_color
red_eyes <- starwars |>
  filter(str_detect(eye_color, "red"))
red_eyes

# Find out which character has red and blue eyes
red_and_blue_eyes <- starwars |> filter(str_detect( eye_color, "blue") & str_detect(eye_color, "red"))
view(red_and_blue_eyes)

#R4-P17 has blue and red eyes
```

## Question 2 — American Unemployment Over Time

The `ggplot2::economics` dataset contains monthly U.S. macroeconomic indicators from 1967 to 2015. The `date` column is already a proper `Date` object.

**(a)** The goal here is to see how the unemployment rate changes over time. Start by plotting the number of unemployed people over time:

```{r}
  ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
```

Now use lubridate to plot only the unemployment data from Jan 1st 2008 to Jan 1st 2009. Hint: use the lubridate function ymd to specify date windows, and use a similar logical criteria to in Q1, above

```{r}
# your code here
economics_2008 <- economics |>
  filter(date > ymd("2008-01-01") & date < ymd("2009-01-01"))

ggplot(economics_2008, aes(x = date, y = unemploy)) +
  geom_line()
```

What do you notice about the format of the date variable on the x axis of this plot vs the plot over the entire time window. What do you think is happening here?

**Your answer: The x-axis of the plot only showing the year 2008-2009 is by month. The x-axis for the plot over the entire time window is by every decade and has less frequency labels. In the 2008-2009 plot, the time is shorter so it has been adjusted to be more specific. I think that ggplot automatically scales the date on the axes based on the range of the dates given.**

Now use mutate to calculate a new variable - the monthly unemployment rate as `unemploy / pop * 100`. Make a plot over the whole time period as in part 1 of this question.

```{r}
# your code here
monthly_unemply_rate <- economics |> mutate(unemployment_rate = unemploy / pop * 100)

ggplot(monthly_unemply_rate, aes(x =  date, y = unemployment_rate)) + geom_line() + labs( title = "Monthly Unemployment Rate Over Time")
```

Do you draw different conclusions now about how unemployment has changed over these 40 years? what explains the difference?

**your answer: I do draw different conclusion. In the original graph, it appeared that unemployment had been increasing over the 40 years. When we switched to unemployment rate you can see that the overall trend is flatter. The increase is the original plot is because the population was increasing. Although the raw number of people unemployed increased, that is simply just because there are more people in the work force, it does not mean that the actual rate increased.**

## Question 3 — Flight Status and the Value of Factors

Run the code below to create a `status` variable that categorises each flight by how early or late it arrived.

```{r}
flight_status <- flights |>
  mutate(
    status = case_when(
      is.na(arr_delay)      ~ "cancelled",
      arr_delay < -30       ~ "very early",
      arr_delay < -15       ~ "early",
      abs(arr_delay) <= 15  ~ "on time",
      arr_delay < 60        ~ "late",
      arr_delay < Inf       ~ "very late",
    ),
    .keep = "used"
  )
```

**(a)** Without converting `status` to a factor, run the followong code to create a bar chart showing the count of flights in each category. What do you notice about the ordering of the bars? Is it meaningful?

```{r}
ggplot(flight_status, aes(x = status)) +
  geom_bar()
```

**your answer: The ordering of the bars is not meaningful, it is just in alphabetical order.**

**(b)** Convert `status` to a factor with levels in a natural severity order: `"very early"`, `"early"`, `"on time"`, `"late"`, `"very late"`, `"cancelled"`. Save it back into `flight_status` and recreate the bar chart. How does the chart change?

```{r}
flight_status <- flight_status |>
  mutate(
    status = factor(
      status,
      levels = c("very early", "early", "on time", "late", "very late", "cancelled")))

ggplot(flight_status, aes(x = status)) +
  geom_bar() 
```

**your answer: The x-axis changes from alphabetic order to natural severity order and is easier to read. It logically makes sense because the first graph is very early, then moves to early, and so on.**
